---
title: "MetabolomicsPipeline Report"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
    code_folding: hide
params: 
  dat: !r MetabolomicsPipeline::demoDat
  includedVarsDist: !r c("GROUP_NAME", "TIME1")
  stratVarDist: !r c("Gender")
  pcaColor: !r c("Gender")
  metsNumber: !r 50
  groupVars: !r c("GROUP_NAME", "TIME1")
  stratVar: !r c("Gender")
  subpath_analysis: !r readRDS("../Data/DemoAnalysisResults/strat_pathway_example.Rds")
  subpath_selection: !r c("Partially Characterized Molecules")
  pairwise_analysis: !r readRDS("../Data/DemoAnalysisResults/pairwise_analysis_example.Rds")
  CHEMICAL_NAME: !r c("CHEMICAL_NAME")
  SUB_PATHWAY: !r c("SUB_PATHWAY")
  subpathway_subpath:  !r c("Lactoyl Amino Acid")
  subpathway_treat: !r c("GROUP_NAME")
  subpathway_block: !r c("TIME1")
  subpathway_filter: !r c("Gender")
  DEMO: !r TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MetabolomicsPipeline)
library(table1)
library(SummarizedExperiment)
library(dplyr)
library(htmltools)
library(ggplot2)
```


# Introduction

This report provides a summary of the metabolomics data analysis using MetabolomicsPipeline dashboard. In this 
report, we include an exploratory data analysis, subpathway analsysis, pairwise analysis
and subpathway plots. The experiment includes a total of `r nrow(colData(params$dat))` samples and `r nrow(rowData(params$dat))` metabolites. If using the MetabolomicsPipeline dashboard for preprocessing, 
the peak data is median standardized, min value imputed, and log transformed for the downstream analysis.

```{r load-data}

# Create table1
form <- as.formula(
    if(params$stratVarDist == "None"){
      paste0("~ ", paste(params$includedVarsDist, collapse = " + "))
    } else {
      paste0("~ ", paste(params$includedVarsDist, collapse = " + "), " | ",
             params$stratVarDist)
    }
    )
  

table1(form, data = colData(params$dat))
```


# Exploratory Data Analysis

## PCA
In general, Principal component analysis (PCA) reduces the number of variables in a dataset while preserving as much information from the data as possible. At a high level, PCA is constructed such that the first principal component (PC) accounts for the largest amount of variance within the data. The second PC accounts for the largest remaining variance, and so on. Additionally, each of the PCs produced by PCA is uncorrelated with the other principal components. PCA can allow us to visualize sources of variation in the data.

```{r pca}
# PCA plot
# Define PCA label from metadata
meta_var <- params$pcaColor

# Run PCA
pca <- metabolite_pca(params$dat,
    meta_var = meta_var
)


# Show PCA
pca
```


## Heatmap

For our heatmap, the x-axis will be the samples, and the y-axis will be the metabolites. The values determining the colors will be the normalized peak values for each metabolite in each observation. We can group the observations by the experimental conditions. Grouping the experimental conditions in a heatmap is another way of visualizing sources of variation.


```{r heatmap, message=FALSE}
metabolite_heatmap(params$dat,
                        top_mets = params$metsNumber,
                        group_vars = params$groupVars,
                        strat_var = params$stratVar,
                        caption = "",
                        Assay = "normalized",
                        !!!syms(params$groupVars))
```


# Subpathway Analysis
Each metabolite is nested within a subpathway, which in turn is nested within a superpathway. To test for differences in peak intensities between treatment groups at the metabolite level, we use an Analysis of Variance (ANOVA) model. However, because multiple metabolites belong to the same subpathway, it is challenging to assess treatment effects at the subpathway level directly. To address this, we apply Fisher’s combined probability test, which aggregates p-values from independent tests to evaluate whether there is an overall effect. This approach allows us to assess subpathway-level effects based on the p-values obtained from metabolite-level models.

### Combined Fished Analysis

We will test at the subpathway level by combining the p-values for each
metabolite within the subpathway for each model. We use a combination
function given by $\tilde{X}$ which combines the pvalues, resulting in a
chi-squared test statistic.

$$
\tilde{X} = -2\sum_{i=1}^k ln(p_i)
$$

where $k$ is the number of metabolites in the subpathway. We can
get a p-value from $P(X \geq\tilde{X})$, knowing that
$\tilde{X}\sim \chi^2_{2k}$. You will notice that smaller p-values will
lead to a larger $\tilde{X}$.

##### Assumptions

Since we are first testing each metabolite utilizing ANOVA, we make the
following assumptions for each metabolite,

-   *Independence:* Each observation is independent of all other
    observations. Therefore, if you have collected multiple samples from
    the same subject then this assumption may be violated.

-   *Normality:* The normalized metabolite intensities follow a normal
    distribution within each of the treatment groups.

-   *Homoscedasticity:* Equal variance between treatment groups.

In addition to the assumptions in the ANOVA models at the metabolite
level, the Fisher's Combined probability places an independence
assumption between the metabolites within the subpathway. 

For more about the Combined Fisher Probability and other methods that
can address this problem, see:

Loughin, Thomas M. "A systematic comparison of methods for combining
p-values from independent tests." Computational statistics & data
analysis 47.3 (2004): 467-485.

#### Models

To test our hypothesis at the subpathway level, we first have to form
our hypothesis at the metabolite level. For each metabolite, we test
three models.

1.) Interaction: $nornalized Peak = Treatment Group + Block + Treatment*Block$

2.) Parallel: $normalized Peak = Treatment Group + Block$

3.) Single: $normalized Peak = Treatment$

For the interaction model, we are focusing only on the interaction term
"Treatment\*Block" to test if there is a significant
interaction between our treatment and the block variable. The parallel
model is testing if the block variable is explaining a significant
amount of the metabolite variance with treatment included, and the treatment
model is testing if the treatment explains a significant proportion of the 
variance for each metabolite.  

```{r subpathway analysis}
subpath_by_model(params$subpath_analysis)

# Create subpath within superpath table
subpath_within_superpath(params$subpath_analysis)

# Show metabolites within subpathway
tables <- met_within_sub(params$subpath_analysis, subpathway = params$subpath_selection)

# Combine and return as HTML
HTML(paste(tables, collapse = "<br><br>"))
```


# Pairwise Analysis
### Log Fold-Change Heatmap
We will produce a heatmap of the log fold changes for the metabolites with a significant overall p-value (which tested if the treatment group means were equal under the null hypothesis). The heatmap colors will only show if the log fold-change is greater than log(2) or less than log(.5). Therefore, this heatmap will only focus on comparisons with a fold change of two or greater.

### P-Value Heatmap
Similar to the pairwise estimate heatmap, we will produce a heatmap where the heatmap will only include metabolites with a significant overall p-value, and the values in the heat map will only be colored if the pairwise comparison is significant. 

```{r est heatmap}
# Read in example pairwise data
pairwise_analysis <- params$pairwise_analysis

# Create estimate heatmap
if(class(pairwise_analysis) == "list"){
  # Get stratums
  stratums <- names(pairwise_analysis)
  # Create heatmap for each stratum
  plots <- lapply(stratums, function(stratum) {
    met_est_heatmap(pairwise_analysis[[stratum]],
                    params$dat,
                    interactive = TRUE,
                    show_rownames = FALSE,
                    plotlyTitle = paste0("Fold Change Heatmap: ", stratum),
                    main = "")
  })
  
  do.call(tagList,plots)
    
} else {
  p <- met_est_heatmap(pairwise_analysis,
                   params$dat,
                   interactive = TRUE,
                   show_rownames = FALSE,
                   plotlyTitle = "Fold Change Heatmap",
                   main = "")
  
  # Print the heatmap
  print(p)
}

```


```{r pvalue heatmap}

# Read in example pairwise data
pairwise_analysis <- params$pairwise_analysis

SUB_PATHAY = params$SUB_PATHWAY

CHEMICAL_NAME = params$CHEMICAL_NAME

# Create estimate heatmap
if(class(pairwise_analysis) == "list"){
  # Get stratums
  stratums <- names(pairwise_analysis)
  # Create heatmap for each stratum
  plots <- lapply(stratums, function(stratum) {
    met_p_heatmap(pairwise_analysis[[stratum]],
                    params$dat,
                    interactive = TRUE,
                    SUB_PATHWAY = SUB_PATHAY,
                    CHEMICAL_NAME = CHEMICAL_NAME,
                    plotlyTitle = paste0("P-value Heatmap: ", stratum))
  })
  
  do.call(tagList,plots)
    
} else {
  p <- met_p_heatmap(pairwise_analysis,
                   params$dat,
                   interactive = TRUE,
                   SUB_PATHWAY = SUB_PATHAY,
                  CHEMICAL_NAME = CHEMICAL_NAME,
                   plotlyTitle = "Pvalue Heatmap")
  
  # Print the heatmap
  print(p)
}

```

# Subpathway Plots
Visualizations of the data can help us see the underlying trends. Two useful visualizations are boxplots and line plots. This allows us to visualize the distribution of the metabolite intensities within a subpathway across treatment groups. 

```{r subpathway boxplots}
dat <- params$dat

# Get the selected values from the inputs
  subpathway_var <- params$SUB_PATHWAY
  subpathway_subpath <- params$subpathway_subpath
  subpathway_treat <- params$subpathway_treat
  subpathway_block <- params$subpathway_block
  subpathway_chemName <- params$CHEMICAL_NAME
  subpathway_filter <- params$subpathway_filter
  
  
  # Create the boxplot
  if(subpathway_filter == "None"){

      subpathway_boxplots(dat,
                    subpathway = subpathway_subpath,
                    block_var = !!sym(subpathway_block),
                    treat_var = !!sym(subpathway_treat),
                    Assay = "normalized",
                    CHEMICAL_NAME = subpathway_chemName,
                    SUB_PATHWAY = subpathway_var) +
      ggtitle(paste0("Boxplots for ", subpathway_subpath))
  } else{
      for (strata in unique(dat[[subpathway_filter]])) {
        
        d_filtered <- dat[,colData(dat)[[subpathway_filter]] == strata ]
        
        p <- subpathway_boxplots(d_filtered,
                    subpathway = subpathway_subpath,
                    block_var = !!sym(subpathway_block),
                    treat_var = !!sym(subpathway_treat),
                    Assay = "normalized",
                    CHEMICAL_NAME = subpathway_chemName,
                    SUB_PATHWAY = subpathway_var) +
      ggtitle(paste0("Boxplots for ", subpathway_subpath), subtitle = strata)
        
        print(p) 
      }
  }
      
      


```

```{r subpathway lineplots}

# Get data
dat <- params$dat

if(params$DEMO==TRUE){
  
  dat$TIME1 <- as.numeric(factor(dat$TIME1,
      levels = c("PreSymp", "Onset", "End")
    ))
}

# Get the selected values from the inputs

  subpathway_var <- params$SUB_PATHWAY
  subpathway_subpath <- params$subpathway_subpath
  subpathway_treat <- params$subpathway_treat
  subpathway_block <- params$subpathway_block
  subpathway_chemName <- params$CHEMICAL_NAME
  subpathway_filter <- params$subpathway_filter

# Create line plots for the subpathway
if(subpathway_filter == "None"){

  subpathway_lineplots(dat,
                    subpathway = subpathway_subpath,
                    block_var = !!sym(subpathway_block),
                    treat_var = !!sym(subpathway_treat),
                    Assay = "normalized",
                    CHEMICAL_NAME = subpathway_chemName,
                    SUB_PATHWAY = subpathway_var) +
  ggtitle(paste0("Lineplots for ", subpathway_subpath))
} else{
  for (strata in unique(dat[[subpathway_filter]])) {
    
    d_filtered <- dat[,colData(dat)[[subpathway_filter]] == strata ]
    
    p <- subpathway_lineplots(d_filtered,
                    subpathway = subpathway_subpath,
                    block_var = !!sym(subpathway_block),
                    treat_var = !!sym(subpathway_treat),
                    Assay = "normalized",
                    CHEMICAL_NAME = subpathway_chemName,
                    SUB_PATHWAY = subpathway_var) +
      ggtitle(paste0("Lineplots for ", subpathway_subpath), subtitle = strata)
    
    print(p) 
  }
}

```

# Session Info
```{r session}
sessionInfo()
```
